Reasonably Programmable Literal Notation

### 2018-03-08 icfp18/syntax-icfp18.tex
near-final revisions up through context dependence section

Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s} and the \emph{type-specific languages} (TSLs) of \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).

@@

\begin{enumerate}[leftmargin=15px]
\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same literal?
\item \textbf{Expansion Typing}: What type does \li{y} have?
\item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on? If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depended invisibly on the original binding?
\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation (e.g. as operators in the K query language)?
\item \textbf{Segment Typing}: What type is each spliced term expected to have? How can I infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 5, or might it capture an invisible binding of the same identifier in the expansion of Line 9?
\end{enumerate}

@@

Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason. We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar. We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}. Briefly, we found that:
\begin{itemize}[leftmargin=15px]
\item Unhygienic approaches, like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms. They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten. However, they do not satisfy the remaining five reasoning criteria.
\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e. the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.  
\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at other types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing. Only existing composite forms can therefore be repurposed.
\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec. \ref{sec:existing-approaches}. In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme. It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
\end{itemize}

@@

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation, of nearly arbitrary design, that (1) provides the ability to reason abstractly about all six of the topics just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages. We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of the claimed abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above. In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger client-side reasoning principles. % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.

@@

\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Figure \ref{fig:K-tsm-example}. On Lines 1-5, we define a TLM named \li{$kq} that provides K query literal notation. On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}. We can reason abstractly about this program as follows:
\begin{enumerate}[leftmargin=12pt]
\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e. the characters between \li{`(} and \li{)`}. We will give more details on generalized literal forms and on constructing a TLM lexer and parser in the next section. For now, let us simply reiterate that our design goal is to provide a system where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding. % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools. %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others. The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site. This completely relieves clients from needing to consider expansion-internal dependencies when naming variables. %For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. The segmentation is all that needs to be communicated to language services downstream of the expander, e.g. editors and pretty printers, which can communicate this information to the programmer using secondary notation, e.g. colors in this document. So by examining Fig. \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators of the K language (because they are green).
\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. by reference only to the type annotations on the spliced segments, not the full expansion.  %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot have captured a coincidental binding of \li{x} in the expansion. % We will say more about capture later on.
\end{enumerate}

@@

\vspace{-3px}
\newcommand{\seTLMsSec}{Expression TLMs}
\section{\protect\seTLMsSec}
\label{sec:setsms}
\vspace{-2px}

Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
The notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}. Even with this shorthand, however, this notation for constructing regexes is rather unwieldy. Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.

% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex. This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}. However, this approach breaks down when one needs to construct regexes compositionally. For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases. However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}. These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}. This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation. When some of these strings come from untrusted sources, e.g. web forms as is common in bioinformatics applications, the result can be a breach in privacy.

@@

\begin{subfigure}[t]{\textwidth}
\vspace{4px}
\begin{lstlisting}
  let any_base = $regex `(SURLA|T|G|CEURL)`;
  let bisA = $regex `(SURLGC$(EURLany_baseSURL)GCEURL)`;
  let gene_restriction_template(gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgeneSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
\end{lstlisting}
\vspace{-5px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{3px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
\vspace{-5px}
\end{figure}

\vspace{-3px}
\subsection{TLM Definition and Application}
Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} for values of type \li{Regex.t} that supports a POSIX-style literal notation extended with splice forms for regexes and strings. Line 1 of Fig. \ref{fig:first-tlm-example} applies this TLM to construct the regex \li{any_base} just described. Line 2 of Fig. \ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splice form to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base. Finally, Lines 3-4 of Fig. \ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using a string splice form delimited by \li{$$(} and \li{)}.

Let us consider the second of these three TLM applications more closely:
\begin{lstlisting}[numbers=none]
    $regex `(GC$(any_base)GC)`
\end{lstlisting}
According to the context-free grammar of the language, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
 to distinguish them from variables. There are no lexical constraints on the body of the generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in OCaml, Reason and other languages. Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, therefore syntactically subsume other literal forms. The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use \li{`(} and \li{)`} exclusively. The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.

Notice that in the unexpanded expression above, we did not distinguish the spliced expression from the rest of the literal. This is to emphasize that it is only during the subsequent \emph{expansion} phase that the body of the generalized literal form is lexed, parsed and expanded to produce an expanded syntax tree where spliced expressions, like \li{any_base}, have been revealed. Responsibility for lexing, parsing and expanding the literal body is delegated to the lexer and parser specified by the applied TLM by the clauses \li{lexer} and \li{parser}, respectively.

\vspace{-4px}
\subsection{Abstract Reasoning Principles} 
Rather than looking immediately at Fig. \ref{fig:lexer-and-parser}, which defines the lexer and parser specified by \li{$regex}, let us first consider the perspective of a client programmer who does not want to delve into the details of lexing and parsing for each piece of literal notation that they encounter. What can this programmer deduce just by reasoning from the information presented in Figures \ref{fig:Regex-module-def} through \ref{fig:first-tlm-example}?

% TLM definitions follow well-defined scoping rules, so the client programmer can follow the binding structure of the language in the usual manner (manually or assisted by a ``go to definition'' editor service) to determine which TLM is uniquely responsible for each literal form. In Sec. \ref{sec:ptsms}, we will see an example of TLM synonyms (as a trivial case of partial parameter application).

From the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig. \ref{fig:regex-tlm-def}. This is analagous to the return type of a function. The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site (so the module \li{Regex} need not be in scope at the application site, or it can be shadowed by a different module).

The TLM mechanism enforces an similarly strong context independence property on generated expansions by requiring that the TLM explicitly specify the modules that the expansions it generates might internally depend on. In this case, Lines 4-5 of Fig. \ref{fig:regex-tlm-def} specify that generated expansion might use the module \li{Regex}, again as it is bound at the definition site, using the module variable \li{Regex} internally. These happen to coincide in this case, but in general, the dependency can be an arbitrary module path while internally it must be bound to a module variable, for example:
\begin{lstlisting}[numbers=none]
  expansions require 
    module Regex as Regex
    module Core.List as L
\end{lstlisting}
The module variable \li{L} will be bound internally to \li{Core.List} in all expansions generated by this TLM, including those where \li{L} or \li{Core.List} are not bound, or otherwise bound. All other bindings, whether at the TLM definition site or the TLM application site, are not internally available to the expansion. This allows client programmers to freely rename and shadow variables as they are used to, without needing to inspect each parser to determine whether it requires certain modules be bound to particular names at each use site (as is common with, e.g., syntax extensions based on \li{camlp4} or Sugar*, or when using unhygienic term rewriting approaches like OCaml's PPX system or Template Haskell \cite{SheardPeytonJones:Haskell-02}, as further discussed in Sec. \ref{sec:existing-approaches}).

Enforcing this strong context independence condition is quite subtle because TLM parsers need to be able to extract spliced expressions from the literal body and place them in the final expansion. Na\"ively checking that the final expansion is closed except for the explicitly named dependencies would inappropriately enforce context independence on application site spliced expressions, which should certainly not be prevented from referring to variables in the application site context. For example, consider the final expansion of the example from Line 2 of Fig. \ref{fig:first-tlm-example} above:
\begin{lstlisting}[numbers=none]
  Regex.Seq(Regex.Str "GC", Regex.Seq(any_base, Regex.Str "GC"))
\end{lstlisting}
In this term, both \li{Regex} and \li{any_base} are free variables. There is nothing to indicate which free variables came from the literal body via an intentional act of splicing by the TLM. 
%\footnote{An exhaustive search would not be feasible because final expansions are abstract syntax trees. There is not a unique mapping back to concrete syntax, e.g. because whitespace can be inserted arbitrarily.} 
As we will discuss more extensively in Sec. \ref{sec:existing-approaches}, this is why traditional hygienic term-rewriting macro systems, like those available in various Lisp-family languages \cite{mccarthy1978history} and in Scala \cite{ScalaMacros2013}, cannot be used to repurpose string literals for literal notation at other types.

To address this problem, the TLM parser does not generate the final expansion directly. Instead, the parser generates a \emph{proto-expansion} that refers to spliced expressions indirectly by location relative to the start of the provided literal body. For example, the proto-expansion generated by \li{$regex} for the example above can be expressed as follows:
\begin{lstlisting}[numbers=none]
  Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4; 11; Regex.t>, Regex.Str "GC"))
\end{lstlisting}

Here, \li{spliced<4; 11; Regex.t>} is a reference to the spliced expression \li{any_base} by its location relative to the start of the literal body being expanded. The context independence condition can be enforced straightforwardly on the proto-expansion -- the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency.
% Requiring that TLMs refer to spliced expressions indirectly in this manner ensures that a TLM cannot ``forge'' spliced terms, i.e. mark some sub-term as being spliced when it does not in fact appear at some specified segment of the literal body.

>>>>

Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).

@@

\begin{enumerate}[leftmargin=15px]
\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same literal?
\item \textbf{Expansion Typing}: What type does \li{y} have?
\item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on? If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depends invisibly on the original binding?
\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation (e.g. as operators in the K query language)?
\item \textbf{Segment Typing}: What type is each spliced term expected to have? How can I infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 5, or might it capture an invisible binding of the same identifier in the expansion of Line 9?
\end{enumerate}

@@

Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason. We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar. We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}. Briefly, we found that:
\begin{itemize}[leftmargin=15px]
\item Unhygienic approaches like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms. They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten. However, they do not satisfy the remaining five reasoning criteria.
\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e. the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.  
\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} (e.g. Racket \cite{Flatt:2012:CLR:2063176.2063195}) and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at composite types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing. Only existing composite forms can therefore be repurposed.
\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec. \ref{sec:existing-approaches}. In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme. It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
\end{itemize}

@@

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provides the ability to reason abstractly about all six of the topics just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages. We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of the claimed abstract reasoning principles; and (2) a number of non-trivial examples appearing throughout the paper that involve the advanced language features mentioned above. In describing these examples, we also demonstrate that literal parsing logic can be defined using standard, unmodified parser generators, so the burden on notation providers is comparable to that of existing systems despite these stronger reasoning principles. % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable.

@@

\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Figure \ref{fig:K-tsm-example}. On Lines 1-5, we define a TLM named \li{$kq} that provides K query literal notation. On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}. We can reason abstractly about this program as follows.
\begin{enumerate}[leftmargin=12pt]
\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e. the characters between \li{`(} and \li{)`}. We will give more details on generalized literal forms and on constructing a TLM lexer and parser in the next section. For now, let us simply reiterate that our design goal is to provide a system where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding. % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools. %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others. The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site. This completely relieves clients from needing to consider expansion-internal dependencies when naming variables. %For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. The segmentation is all that needs to be communicated to language services downstream of the expander, e.g. editors and pretty printers, which can communicate this information to the programmer using secondary notation, e.g. colors in this document. So by examining Fig. \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators of the K language (because they are green).
\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. by reference only to the type annotations on the spliced segments, not the full expansion.  %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot have captured a coincidental binding of \li{x} in the expansion. % We will say more about capture later on.
\end{enumerate}

@@

\vspace{-3px}
\newcommand{\seTLMsSec}{Expression TLMs in Reason}
\section{\protect\seTLMsSec}
\label{sec:setsms}
\vspace{-2px}

Consider the recursive datatype \li{Regex.t} defined by Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we did not need to qualify each application of \li{Regex.Or} and \li{Regex.Str} above. Even with this shorthand, however, constructing regexes in this way is syntactically costly. Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from previously constructed regexes and strings. 
We solve this problem in Fig. \ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings. 
% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex. This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}. However, this approach breaks down when one needs to construct regexes compositionally. For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases. However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}. These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}. This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation. When some of these strings come from untrusted sources, e.g. web forms as is common in bioinformatics applications, the result can be a breach in privacy.

@@

\begin{subfigure}[t]{\textwidth}
\vspace{4px}
\begin{lstlisting}
  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
  let restriction_template(gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
\end{lstlisting}
\vspace{-2px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{3px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
\vspace{-5px}
\end{figure}

\vspace{-3px}
\subsection{Client Perspective}
The focus of this paper is on the client programmer's perspective, so we start in this section with some examples of \li{$regex} being applied in Fig. \ref{fig:first-tlm-example}, then use these examples to detail the client-side abstract reasoning principles that the mechanism provides, all before finally considering the details of the lexer and parser in Sec. \ref{sec:provider-perspective}. % The \li{open notation} directive on Line 2 of Fig. \ref{fig:first-tlm-example} allow qualifying each TLM application in the file. %reasoning about this program from the perspective of a client who wants to avoid looking at the lexer, parser and the underlying expansion itself.

Line 1 of Fig. \ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{any_base} that was  described above, this time using POSIX regex notation. Line~2 applies \li{$regex} again, using its regex splice form to construct a regex matching DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base. Finally, Lines~3-4 of Fig. \ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using the string splice form. Notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form.

Let us consider the second of these three TLM applications more closely:
\begin{lstlisting}[numbers=none]
    $regex `(GC$(DNA.any_base)GC)`
\end{lstlisting}
According to the context-free grammar of (our extended dialect of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
 to distinguish them from variables. We call the TLM argument, \li{`(GC$(DNA.any_base)GC)`}, a \emph{generalized literal form}, following the terminology introduced in the prior work on type-specific languages \cite{TSLs}. The only lexical constraint on the body of a generalized literal form, i.e. the characters between \li{`(} and \li{)`}, is that any further occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml. Generalized literal forms therefore lexically subsume other literal forms. The nesting constraint is to allow TLM applications to appear inside spliced expressions (discussed below). The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use the outer delimiters \li{`(} and \li{)`} exclusively. Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM, as we will discuss below.

% Notice that in the unexpanded expression, we have not yet distinguished the spliced expression \li{any_base} from the rest of the literal. This is to emphasize that it is only during the subsequent \emph{expansion} phase that the body of the generalized literal form is lexed, parsed and expanded to produce an expanded syntax tree where spliced expressions, like \li{any_base}, have been revealed.

% \vspace{-4px}
% \subsection{Client Perspective: Abstract Reasoning Principles} 
% The question that this paper posits is central when evaluating a notation definition system is: what must be revealed about the lexer, parser and the expansion itself for the client programmer to be able to answer the sorts of questions about types and binding that they would normally ask?

\paragraph{Responsibility} 
Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the lexer and parser specified in the definition of the applied TLM by the clauses \li{lexer} and \li{parser}, respectively. We will return to consider the details of lexer and parser definition, and show the definitions of the lexer and parser specified by \li{$regex}, in Sec. \ref{sec:provider-perspective}. For now, let us purposefully hold these details abstract.% and take the perspective of a client that does not want to delve into these details.

In Reason, TLM definitions like the one in Fig. \ref{fig:regex-tlm-def} are separately defined in files with the suffix \li{.relit}, which are processed and brought into scope by the Reason build system, \li{rebuild}. This is consistent with the protocol that other syntax extension systems use, e.g. the syntax libraries in Sugar* \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}. In our theoretical treatment in Sec.~\ref{sec:setsms-formally}, however, TLM definitions are integrated directly into the scoping structure of the language. This is not possible in Reason because this would require extending OCaml's semantics, i.e. adding TLM definitions to module signatures. While we were careful to choose a design where this modification would be feasible in principle---TLM definition equality requires only decidable type and module path equality---modifying the semantics of OCaml is intentionally beyond the scope of the Reason project. In any case, what is fundamentally important is that there is a well-defined protocol for finding the definition of the TLM uniquely responsible for each application site. % via a ``go to definition'' service.

% nitions (Fig. \ref{fig:regex-tlm-def}) appear in \li{.relit} files, and they are brought into scope by the build system, \li{rebuild}, according to the usual protocols for library discovery. The fully qualified name of the \li{--}% In the theory that we develop in Sections \ref{sec:s-UL}, TLM definitions occur directly within code. However, in a language with a module system, this would require adding TLM definitions to signatures as well. 
% TLM definitions follow well-defined scoping rules, so the client programmer can follow the binding structure of the language in the usual manner (manually or assisted by a ``go to definition'' editor service) to determine which TLM is uniquely responsible for each literal form. In Sec. \ref{sec:ptsms}, we will see an example of TLM synonyms (as a trivial case of partial parameter application).

\paragraph{Expansion Typing}
Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line 1 of Fig. \ref{fig:regex-tlm-def}. The expansion type of a TLM is analagous to the return type of a function. The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module. In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e. other top-level modules in the package as well as the modules exported by other packages that have been imported by the build system.
